---
title: "Assignment 3: Data Exploration"
author: "Sam Saltman, Section #2"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}
getwd() #check working directory
library(tidyverse) # load necessary package

Neonics <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE) 
Litter <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)

```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: We may want to understand how this type of insecticide affects insects. Insects can harm crops; however, some insects can be good for agricultural purposes by pollinating crops and/or preying on harmful insects in the field. Data analysis on neonicotinoids can help us understand if an insecticide is successful at targeting harmful insects while not causing harms to insects that are beneficial to the agricultural process. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: We may want to study the litter and woody debris in forests to determine the health of the forest. This kind of debris will decompose and provide nutrients to the soil. Litter will decompose faster than woody debris. Analyzing litter and woody debris can  help researchers understand the replenishment rate of nutrients to the soil and the general lifecycle of a forest.

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
* The woody debris and litter samples were collected using both elevated and ground traps. 
* The research used a spatial sampling design to randomly select plot sites. The sites had to contain woody vegetation taller than 2 meters.
* Also, the research had a temporal sampling design. Ground traps were measured once a year. Sampling of elevated traps occurred frequently throughout the year but varied depending on the vegetation in the sample.


## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics) # dimensions
```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
Neonics_effects <- summary(Neonics$Effect)
Neonics_effects
```

> Answer: Population (1803) & Mortality (1493) are the most common effects studied. Studying population may be of interest to see if the insecticide increases, decreases or has no effect on a population size over the course of the study. Studying mortality could help researchers understand the effectiveness of the insecticide as well as understand why population size may have changed.

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name)
```

> Answer: Top 6: Honey Bees, Parasitic Wasp, Buff Tailed Bumblebee,Carniolan Honey Bee, Bumble Bee, Italina Bee. Bees are the most studied because a healthy garden needs pollinators. Researchers want to make sure that these insects are not harmed by the insecticide. The Asian Lady Beetle is also frequently studied because it is beneficial to a farm by preying on harmful insects. The insecticide should not harm this species of insect either. On the other hand, the Japanese Beetle is studied frequently as they are a pest. The insecticide should be effective at reducing this insect population.

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.) #it's a factor

```

> Answer: It is not numeric because it contains both numbers and characters.



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 50)
  
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 50) 
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Labs and field natural test locations are most commonly used. There is variability over time, but it appears that these locations have been used more often since the first recorded publish date in this dataset. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}

ggplot((Neonics)) +
  geom_bar(aes(x = Endpoint)) +
  theme(axis.text.x = element_text(angle = 90)) # used this code to make x axis labels visible

```

> Answer: LOEL is the "lowest-observable-effect-level," and it is defined as the lowest dose that would produce a significantly  different response from the controls. 
> NOEL is "no-observable-effect-level," and it is defined as the highest dose that does not cause a significantly different response from the controls.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate) #it's a factor to start. Can run this code again to see if conversion worked below as well
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d") #change factor to date
class(Litter$collectDate)
unique(Litter$collectDate) #samples were collected on August 2nd and 30th in 2018

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID)
```

> Answer: The unique function lists every unique plot ID and gives the total of the different sites which is 12 in this case. We can infer the same info from the summary function; however, this function does not explicitly state how many unique plots there are. It's main function is to provide stats on each plot.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
ggplot(Litter) +
  geom_bar(aes(x = functionalGroup))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass))

ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot helps us visualize the range of drymass between functionalgroups. We can clearly see the calculated averages between the samples. However, drymass accumulation in all functionalGroups varies between each sample. The violin is good for visually communicating if there is a certain range with many data points. This isn't the case in this chart, so the violin becomes less informative than the boxplot in this instance as a result.

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles