---
title: "Assignment 2: Coding Basics"
author: "Sam Saltman"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
seq(1, 100, 4) #1 = lower range, 100 = upper range, 4 = increasing factor

#2. 
mean(seq(1, 100, 4)) #Determine mean
median(seq(1, 100, 4)) #Determine median
summary(seq(1, 100, 4)) #Finds statistics including mean and median
#3. 
mean(seq(1, 100, 4)) > median(seq(1, 100, 4)) #False as mean = median
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
Names.of.students <- c("Sam", "Michael", "Tom", "Jackson")
class(Names.of.students) #Character vector

Test.Scores <- c(100, 80, 60, 40)
class(Test.Scores) #Numeric vector

PassingScore <- c("TRUE", "TRUE", "TRUE", "FALSE")
class(PassingScore) #Character vector

Class.Test.Scores <- data.frame("Name"=Names.of.students,"Score"=Test.Scores,"Pass"=PassingScore)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A data frame can store multiple data types. The dataframe created above has both numeric and charactor data within the columns. A matrix is different because it can only store one type of data between columns.

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.
```{r}
Pass.Fail <- function(x){
  ifelse(x>=50, print(TRUE), print(FALSE))
}

```

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
Pass.Fail(Test.Scores) #Ifelse method



```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
